
# Backgrounds {#backgrounds}



```{r child="_includes/_02_01-historical_background.Rmd"}


```


```{r child="_includes/_02_02-modern_technological_approaches.Rmd"}


```

Having motivated this work in terms of historical context up to contemporary technological approaches to improving playing by ear skills, I will now turn to motivating it in terms of the more specific epistemological approaches taken to assess this question in this dissertation. These approaches are: systematic musicological, cognitive psychological, and computational. In general, these approaches come from the *scientific approach*, whereas the content reviewed in the last chapter was largely not based on such an approach.

```{r child="_includes/_02_03-playing_by_ear_literature_review.Rmd"}


```


```{r child="_includes/_02_04-musicological_background.Rmd"}


```

```{r child="_includes/_02_05-cognitive_psychological_background.Rmd"}


```

```{r child="_includes/_02_06-statistical_modelling_frameworks.Rmd"}


```

