

# Experiment 5: A Similarity-Based Approach to Item Selection {#similarity_study}


## Collaborative Filtering




## Approaches to Similarity for Large-Scale Corpora


```{r, include = FALSE}

# Problem:

# WJD_no_items <- nrow(WJD::WJD('phrases'))

```

```{r, include = FALSE}

# factorial(WJD_no_items)

```


## Melodic corpora as networks 




### Approach #1: Similarity of Features as a Proxy for Melodic Similarity

Conclusion: similarity of features =! melodic (opti3/[perceptual?!]) similarity
- this is important, because similar features will produce similar difficulty scores (need to test this formally though).. which shows that similarity as a separate item selector to difficulty is warranted/important (potentially)


### Approach #2: Generative Similarity

- bringing in desimilarize et al. app


### Approach #3: Parent Melody Representation

- On breaking up items into N-grams, store representations/similarity.
- By definition N-grams will be similarity to the overall melody they come from

